Time series clusters

### 2020-06-19 src/ts_clusters.Rmd

---
title: "Time series plots"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

```{r settings, include = FALSE}
library(tsbox)
library(ggplot2)
library(tidyverse)
library(geojsonio)
library(tsfeatures)
library(rprojroot)
library(openair)
library(reshape2)
library(lubridate)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(classInt)
library(RColorBrewer)
library(knitr)
library(rprojroot)
library(viridis)
library(hrbrthemes)

@@

# Structured TS

@@

# map LA mean test frequency per date
la@data <- left_join(la@data, la.dateAvg, by = "LAD19NM")
var <- la@data[ ,'mean(n)']
breaks <- classIntervals(var, n = 5, style = "quantile")
my_colours <- brewer.pal(5, "Blues")
plot(la, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
                 border = rgb(0.8, 0.8, 0.8, 0))
# drop LA
rm(la)
```

## Sub-daily timeseries

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=500, fig.width=12}

# create sub daily blocks
bb2020sp$block[bb2020sp$hour>=0 & bb2020sp$hour<8] <- 1
bb2020sp$block[bb2020sp$hour>=8 & bb2020sp$hour<16] <- 2
bb2020sp$block[bb2020sp$hour>=16 & bb2020sp$hour<24] <- 3

#bb2020sp$date.block <- paste(bb2020sp$date, bb2020sp$date.block, sep = "_")
# I use block as hour for the ts_wide()
bb2020sp$date.block = ymd_h(paste(bb2020sp$date, bb2020sp$block))
head(bb2020sp)

ts <- bb2020sp %>%
  group_by(date.block,LAD19NM) %>%
  summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())
sapply(ts, function(x) sum(is.na(x)))

#hist(ts$n.tests)
#summary(ts$n.tests)
# I am keeping all the obs. despite the n. of tests for now

# spagetti plot
ggplot(ts[ts$LAD19NM=="City of London",], 
       aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = LAD19NM)) +
  geom_line() + guides(colour=FALSE) + ggtitle("City of London") + #xlab("") +
  ylab("download speed") +
  #geom_text_repel(aes(label=outlier.io), cex = 4) + #this line is from the previous version
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5))

for (i in ts[,2]){
  spaghetti <- ggplot(ts[ts$LAD19NM==i,], 
       aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = LAD19NM)) +
    geom_line() + guides(colour=FALSE) + ggtitle("City of London") + #xlab("") +
    ylab("download speed") +
    #geom_text_repel(aes(label=outlier.io), cex = 4) + #this line is from the previous version
    scale_y_continuous(labels = scales::comma) +
    scale_x_datetime(date_breaks = "1 month") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
    plot.title = element_text(hjust = 0.5))
  print(spaghetti)
}

# https://www.data-to-viz.com/caveat/spaghetti.html
tmp <- ts %>%
  mutate(LAD19NM2=LAD19NM)

tmp %>%
  ggplot( aes(x=as.POSIXct(date.block), y=mean.down)) +
    geom_line( data=tmp %>% dplyr::select(-LAD19NM), aes(group=LAD19NM2), color="grey", size=0.5, alpha=0.5) +
    geom_line( aes(color=LAD19NM), color="#69b3a2", size=1.2 )+
    scale_color_viridis(discrete = TRUE) +
    theme_ipsum() +
    theme(
      legend.position="none",
      plot.title = element_text(size=14),
      panel.grid = element_blank()
    ) +
    ggtitle("A spaghetti chart of baby names popularity") +
    facet_wrap(~LAD19NM)

ggplot(subset(ts,date.block>"2019-12-31"), aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = "green")) +
  geom_line() + guides(colour=FALSE) + ggtitle("Download speeds") + #xlab("") +
  stat_summary(fun.y=mean, geom="line", colour="grey", aes(group=1)) + 
  ylab("download speed") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  #geom_vline(xintercept = as.numeric(as.Date("2020-03-01"))) +#, linetype="dashed", color = "blue", size=1.5) +
  facet_wrap(~LAD19NM, ncol = 2)

@@

>>>>

---
title: "Time series clusters"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output")
  })
---

```{r settings, include = FALSE}
library(tsbox)
library(ggplot2)
library(tidyverse)
library(geojsonio)
library(tsfeatures)
library(rprojroot)
library(openair)
library(reshape2)
library(lubridate)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(classInt)
library(RColorBrewer)
library(knitr)
library(rprojroot)
library(viridis)
library(hrbrthemes)
library(imputeTS)
library(tsbox)
library(dtwclust)

@@

## Structured time series

```{r eval=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}

@@

# map LA mean test frequency per date
la@data <- left_join(la@data, la.dateAvg, by = "LAD19NM")
var <- la@data[ ,'mean(n)']
breaks <- classIntervals(var, n = 5, style = "quantile")
my_colours <- brewer.pal(5, "Blues")
plot(la, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
                 border = rgb(0.8, 0.8, 0.8, 0))

@@

## Keep working hours and days

The time series are based on mean upload and download speeds as well as the total number of tests for working days and during working hours

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}

bb2020sp <- bb2020sp %>% arrange(datetext)

# keep working hours
bb2020sp <- bb2020sp[bb2020sp$hour>=9 & bb2020sp$hour<17,]

# drop w/e
bb2020sp <- bb2020sp %>%
  filter(bb2020sp$weekday!="Sat" & bb2020sp$weekday!="Sun")

head(bb2020sp)
```

## Time series clusters

I create clusters based on upload, download and number of tests time series for tests working hours and days from 16/2/2020 onwards.

More work needs to be done to select the optimal *k* as well as the clustering methods.
The `dtwclust` package comes with quite some readings, saved in the `.\literature\`.

Also, we need to think about upload / download / differnces / period + **imputation**.

# download
ts.down <- bb2020sp %>%
  group_by(date,LAD19NM) %>%
  summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())

# upload
ts.up <- bb2020sp %>%
  group_by(date,LAD19NM) %>%
  summarise(mean.up = mean(speedup))

# tests
ts.tests <- bb2020sp %>%
  group_by(date,LAD19NM) %>%
  summarise(n.tests = n())
```

### Upload speed

```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}

# upload working hours and days

ts.wide <- ts.up %>%
  select(date, LAD19NM, mean.up) %>%
  filter(date > "2020-2-15")

# turn to a wide format
ts.wide <- ts_wide(ts.wide)

# impute with means
library(imputeTS)
ts.wide <- na_mean(ts.wide)
#sapply(ts.wide, function(x) sum(is.na(x)))

# extract names
ts.wide.names <- names(ts.wide[,-1])
ts.wide <- ts.wide[,-1]

# standardise
ts.wide <- zscore(ts.wide)

# convert to list for tsclust
ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))

# 5 - 20 cluster sollutions
pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
                type="partitional",
                distance = "dtw_basic", centroid = "pam", trace = T, 
                args = tsclust_args(dist = list(window.size = 20L)))

# decide k, see RJ-2019-023 for max/min
names(pc_k) <- paste0("k_", c("5","10","15", "20"))
sapply(pc_k, cvi, type = "internal")

# I need to work on the selection

pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
                type="partitional", 
                distance = "dtw_basic", centroid = "pam", trace = T, 
                args = tsclust_args(dist = list(window.size = 20L)))

plot(pc_k)

cluster_up <- data.frame(LAD=ts.wide.names,
                           cluster=pc_k@cluster)
```

### Download speed

```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}

# downlad working hours and days

ts.wide <- ts.down %>%
  select(date, LAD19NM, mean.down) %>%
  filter(date > "2020-2-15")

# turn to a wide format
ts.wide <- ts_wide(ts.wide)

# impute with means
library(imputeTS)
ts.wide <- na_mean(ts.wide)
#sapply(ts.wide, function(x) sum(is.na(x)))

# extract names
ts.wide.names <- names(ts.wide[,-1])
ts.wide <- ts.wide[,-1]

# standardise
ts.wide <- zscore(ts.wide)

# convert to list for tsclust
ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))

# 5 - 20 cluster sollutions
pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
                type="partitional",
                distance = "dtw_basic", centroid = "pam", trace = T, 
                args = tsclust_args(dist = list(window.size = 20L)))

# decide k, see RJ-2019-023 for max/min
names(pc_k) <- paste0("k_", c("5","10","15", "20"))
sapply(pc_k, cvi, type = "internal")

# I need to work on the selection

pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
                type="partitional", 
                distance = "dtw_basic", centroid = "pam", trace = T, 
                args = tsclust_args(dist = list(window.size = 20L)))

plot(pc_k)

cluster_down <- data.frame(LAD=ts.wide.names,
                         cluster=pc_k@cluster)
```

### N. of speeds

```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}

# tests working hours and days

ts.wide <- ts.tests %>%
  select(date, LAD19NM, n.tests) %>%
  filter(date > "2020-2-15")

# turn to a wide format
ts.wide <- ts_wide(ts.wide)

# impute with means
library(imputeTS)
ts.wide <- na_mean(ts.wide)
#sapply(ts.wide, function(x) sum(is.na(x)))

# extract names
ts.wide.names <- names(ts.wide[,-1])
ts.wide <- ts.wide[,-1]

# standardise
ts.wide <- zscore(ts.wide)

# convert to list for tsclust
ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))

# 5 - 20 cluster sollutions
pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
                type="partitional",
                distance = "dtw_basic", centroid = "pam", trace = T, 
                args = tsclust_args(dist = list(window.size = 20L)))

# decide k, see RJ-2019-023 for max/min
names(pc_k) <- paste0("k_", c("5","10","15", "20"))
sapply(pc_k, cvi, type = "internal")

# I need to work on the selection

pc_k <- tsclust(ts.wide, k = 15, seed = 94L, 
                type="partitional", 
                distance = "dtw_basic", centroid = "pam", trace = T, 
                args = tsclust_args(dist = list(window.size = 20L)))

plot(pc_k)

cluster_tests <- data.frame(LAD=ts.wide.names,
                           cluster=pc_k@cluster)

clusters <- merge(cluster_up, cluster_down, by = "LAD")
names(clusters)[2] <- "cluster.up"
names(clusters)[3] <- "cluster.down"
clusters <- merge(clusters, cluster_tests, by = "LAD")
names(clusters)[4] <- "cluster.tests"
#sapply(clusters, function(x) sum(is.na(x)))

clusters  %>% arrange(cluster.up) %>% kable() 
```

## Mapping the clusters

la@data <- merge(la@data, clusters, by.x = "LAD19NM", by.y = "LAD")

var <- la@data[ ,'cluster.up']
plot(la, col = var, main = "Upload clusters",
     border = rgb(0.8, 0.8, 0.8, 0))

var <- la@data[ ,'cluster.down']
plot(la, col = var, main = "Download clusters",
     border = rgb(0.8, 0.8, 0.8, 0))

var <- la@data[ ,'cluster.tests']
plot(la, col = var, main = "Tests clusters",
     border = rgb(0.8, 0.8, 0.8, 0))
